---
title: "preliminary_data_cleaning"
output: github_document
always_allow_html: true
---

```{r setup, include=FALSE}
library(haven)
library(tidyverse)
library(plotly)
```

## Importing NHATS data and selecting only relevant variables

Using only round 5 data, included ID variable, the two variables needed for the depression risk scale (PHQ-2), gender, age, race, whether participant uses mobility devices, size of social network, whether or not they have no one to talk to, self reported memory issues causing problems, self-rated health, and whether they are happy with their living situation. 

In creating dichotomous variable phq_cat regarding depression status, phq.total was subtracted by 2 to account for the fact that total scores on the PHQ-2 should range from 0-6 (sum of two items ranging from 0-3, while scores for the two items provided in our data ranged from 1-4). Once 2 was subtracted from the total score, phq_cat was dichotomized based on a cutoff point of 3 or higher.

Categorical variables were recoded to include descriptions of values and converted into factors.

The following variables will be included in analyses:

* `gender`: Gender (Male, Female)
* `age_cat`: Categorical age at interview (65 to 69, 70 to 74, 75 to 79, 80-84, 85-89, 90+)
* `race_ethnicity`: Race and hispanic ethnicity (White, non-hispanic; Black, non-hispanic; Other, non-Hispanic; Hispanic)
* `cane_walker_wheelchair`: Used cane, walker, wheelchair (Yes, No)
* `memory`: Rate your memory (Excellent, Very Good, Good, Fair, Poor)
* `overall_health`: Overall health condition (Excellent, Very Good, Good, Fair, Poor)
* `likes_living_situation`: Likes living situation (Agree a lot, Agree a little, Agree not at all)

The outcome variable of interest is:

*`PHQ-cat`: Depression (dichotomized based on total PHQ-2 score)

```{r nhats}
nhats = read_sas("./final_data/nhats_r5.sas7bdat")  %>%
  select(spid, hc5depresan1, hc5depresan2, r5dgender, r5d2intvrage, rl5dracehisp, md5canewlker, 
         sn5dnumsn, cg5ratememry, hc5health, wb5truestme4) %>%
  janitor::clean_names() %>%   
  filter(hc5depresan1 > 0, hc5depresan2 > 0, sn5dnumsn >= 0, r5d2intvrage > 0, md5canewlker > 0, 
         cg5ratememry > 0, hc5health > 0, wb5truestme4 > 0, rl5dracehisp < 5)

nhats %>% 
  select(hc5depresan1, hc5depresan2)  %>% 
  rowSums(na.rm = TRUE) -> nhats$phq.total

nhats = nhats %>%
  mutate(phq.total = phq.total - 2) %>% 
  mutate(phq_cat = if_else(phq.total < 3, "Mild/No Depression", "Moderate/Severe Depression")) %>% 
  mutate(
    gender = recode(r5dgender, `1` = "Male", `2` = "Female"),
    gender = factor(gender),
    age_cat = recode(r5d2intvrage, `1` = "65 to 69", `2` = "70 to 74", `3` = "75 to 79", 
                 `4` = "80 to 84", `5` = "85 to 89", `6` = "90+"),
    age_cat = factor(age_cat, levels = c("65 to 69", "70 to 74", "75 to 79", "80 to 84", 
                  "85 to 89", "90+"), ordered = TRUE),
    race_ethnicity = recode(rl5dracehisp, `1` = "White, non-hispanic", `2` = "Black, non-hispanic", 
                  `3` = "Other, non-hispanic", `4` = "Hispanic"),
    race_ethnicity = factor(race_ethnicity),
    cane_walker_wheelchair = recode(md5canewlker, `1` = "Yes", `2` = "No"),
    cane_walker_wheelchair = factor(cane_walker_wheelchair),
    memory = recode(cg5ratememry, `1` = "Excellent", `2` = "Very Good", `3` = "Good", 
                    `4` = "Fair", `5` = "Poor"),
    memory = factor(memory, c("Excellent", "Very Good", "Good", "Fair", "Poor"), 
                    ordered = TRUE),
    overall_health = recode(hc5health, `1` = "Excellent", `2` = "Very Good", `3` = "Good", 
                    `4` = "Fair", `5` = "Poor"),
    overall_health = factor(overall_health, levels = c("Excellent", "Very Good", "Good", "Fair", 
                    "Poor"), ordered = TRUE),
    likes_living_situation = recode(wb5truestme4, `1` = "Agree a lot", `2` = "Agree a little", 
                    `3` = "Agree not at all"),
    likes_living_situation = factor(likes_living_situation, c("Agree a lot", "Agree a little", 
                    "Agree not at all"), ordered = TRUE)
)
```

Preliminary exploratory visualizations: 
# for gender - 1 male, 2 female 
# for age categories - 1 65-69, 2 70-74, 5 75-79, 4 80 - 84, 5 85-89, 6 90+
# for racehisp  - 1 white non-hispanic, 2 black non-hispanic, 3 other non-hispanic, 4, hispanic, 5 more than one/dkrf primary

```{r nhats_demo}

gender_bar = 
  nhats  %>%   
  count(gender) %>% 
  mutate(gender = as.factor(gender),
         gender = fct_reorder(gender, n)) %>% 
  plot_ly(x = ~gender, y = ~n, color = ~gender, type = "bar", colors = "viridis") %>% 
  layout(plot_bgcolor = "f8f8f8",
         xaxis = list(title = "participant gender", tickfont = list(size = 11)), 
         yaxis = list(title = "frequency"), 
         title = "Gender",  
         titlefont = list(size = 16),
         showlegend = FALSE)

age_bar = 
  nhats  %>%   
  count(age_cat) %>% 
  mutate(age_cat = as.factor(age_cat),
         age_cat = fct_reorder(age_cat, n)) %>% 
  plot_ly(x = ~age_cat, y = ~n, color = ~age_cat, type = "bar", colors = "viridis") %>% 
  layout(plot_bgcolor = "f8f8f8",
         xaxis = list(title = "participant age", tickfont = list(size = 11)), 
         yaxis = list(title = "frequency"), 
         title = "Age",  
         titlefont = list(size = 16),
         showlegend = FALSE)

race_bar = 
  nhats  %>%   
  count(race_ethnicity) %>% 
  mutate(race_ethnicity = as.factor(race_ethnicity),
         race_ethnicity = fct_reorder(race_ethnicity, n)) %>% 
  plot_ly(x = ~race_ethnicity, y = ~n, color = ~race_ethnicity, type = "bar", colors = "viridis") %>% 
  layout(plot_bgcolor = "f8f8f8",
         xaxis = list(title = "participant race/ethnicity", tickfont = list(size = 11)), 
         yaxis = list(title = "frequency"), 
         title = "Race/Ethnicity",  
         titlefont = list(size = 16),
         showlegend = FALSE)

demo_bar = subplot(gender_bar, age_bar, race_bar, nrows = 3) %>%
  layout(xaxis3 = list(title = "categories"), 
         yaxis2 = list(title = "frequency"),
         title = list(text = "Demographic frequencies"), 
         titlefont = list(size = 14),
         legend = list(font = list(size = 10)))

demo_bar 

```

Plotting of likes living situation predictor against PHQ-2, stratified by race (EM)
#currently broken :( can anyone tell why?

```{r nhats_living_situation}

ll_white = 
  nhats %>%   
  filter(rl5dracehisp == 1 ) %>% 
  plot_ly(x = ~jitter(wb5truestme4, 1), y = ~jitter(phq.total, 1), type = "scatter", 
          mode = "markers", marker = list(color = "red")) %>% 
  layout(plot_bgcolor = "fffefa") 

ll_black = 
  nhats %>%   
  filter(rl5dracehisp == 2 ) %>% 
  plot_ly(x = ~jitter(wb5truestme4, 1), y = ~jitter(phq.total, 1), type = "scatter", 
          mode = "markers", marker = list(color = "blue")) %>% 
  layout(plot_bgcolor = "fffefa") 

ll_other = 
  nhats %>%   
  filter(rl5dracehisp == 3 ) %>% 
  plot_ly(x = ~jitter(wb5truestme4, 1), y = ~jitter(phq.total, 1),  type = "scatter", 
          mode = "markers", marker = list(color = "green")) %>% 
  layout(plot_bgcolor = "fffefa") 

ll_hisp = 
  nhats %>%   
  filter(rl5dracehisp == 4 ) %>% 
  plot_ly(x = c~jitter(wb5truestme4, 1), y = ~jitter(phq.total, 1), type = "scatter", 
          mode = "markers", marker = list(color = "orange")) %>% 
  layout(plot_bgcolor = "fffefa") 


ll_scatter = subplot(ll_white, ll_black, ll_other, ll_hisp, nrows = 4, shareX = TRUE, shareY = TRUE) %>%
  layout(xaxis = list(
    title = "likes living situation", 
    nticks = 3), 
    ticktext = list("Agree a lot", "Agree a little", "Agree not at all"),
    tickvals = list(1, 2, 3), 
         yaxis = list(title = "PHQ-2"),
         title = list(text = "Participant view of living situation and PHQ-2 score distribution"), 
         titlefont = list(size = 14))
         
ll_scatter 

```

Plotting social network and health by phq_cat:

```{r}
sn_phq = nhats %>% 
  ggplot(aes(x = sn5dnumsn, group = phq_cat)) + 
    geom_bar(aes(y = ..prop.., fill = factor(..x..)), stat = "count") +
    geom_text(aes( label = scales::percent(round(..prop.., 3)),
                   y = ..prop.. ), stat = "count", vjust = -.3, size = 2.5) +
    labs(y = "Percent", fill = "SN") +
    facet_grid(~phq_cat) +
    scale_y_continuous(labels = scales::percent)

ggplotly(sn_phq)

health_phq = nhats %>% 
  filter(hc5health > 0) %>%
  ggplot(aes(x = hc5health, group = phq_cat)) + 
    geom_bar(aes(y = ..prop.., fill = factor(..x..)), stat = "count") +
    geom_text(aes( label = scales::percent(round(..prop.., 3)),
                   y = ..prop.. ), stat = "count", vjust = -.3, size = 2.5) +
    labs(y = "Percent", fill = "SN") +
    facet_grid(~phq_cat) +
    scale_y_continuous(labels = scales::percent)

ggplotly(health_phq)
```


####  Visualizing any differences in depression outcome by predictor, self-rated memory. 

- Survey respondents were asked to rate their memory on a scale of 1-5, 1 being Excellent and 5 being Poor. This plot is a visualiztion of the proportion of respondents in this dataset by their depression outcome status - Mild/No Depression or Moderate/Severe Depression. The figure is broken down by how participants rated their memory (x-axis). 

```{r}
mem_plot = nhats %>%  group_by(phq_cat, memory) %>% 
  summarise(n = n()) %>% mutate(
  proportion = (n / sum(n)))

  memory_plot = mem_plot %>% mutate(memory = factor(memory)) %>% drop_na() %>%   
    ggplot(aes(x = memory, y = proportion, group= phq_cat)) + geom_line(alpha = 0.4) +
  
    geom_point(alpha = 0.5, color = "red") + 
    facet_grid(~phq_cat) + 
    labs( title = "Self rated memory and depression outcomes",
    x = "Self rated memory", y = " Proportion of respondents")+  
    scale_y_continuous(labels = scales::percent)
  
ggplotly(memory_plot)
```


####  Visualizing any differences in depression outcome by predictor, use of cane walker wheelchair. 

- Survey respondents were asked if they used a cane walker wheelchair. This plot is a visualization of the proportion of respondents in this dataset by their depression outcome status - Mild/No Depression or Moderate/Severe Depression. The figure is broken down by if participants responded Yes, or No (x-axis). 
```{r}
cane_walker_wc_plot = nhats %>%  
    group_by(phq_cat, cane_walker_wheelchair) %>% 
    summarise(n = n()) %>%  
    mutate(proportion = (n / sum(n)))

cane_walker_wc_plot = 
  cane_walker_wc_plot %>%  
  ggplot(aes(x = cane_walker_wheelchair, y =proportion)) + 
  geom_col(fill= "light blue") + facet_grid(~phq_cat)+
  labs( title = "Use of cane walker wheelchair and depression outcomes",
      x = "Asked if used a cane walker wheel chair", y = " Proportion of respondents")+  
    scale_x_discrete(
      labels = c("Yes","No")) + 
    scale_y_continuous(labels = scales::percent)

  ggplotly(cane_walker_wc_plot)
```

